Parameter Inference for a Simple Stochastic Hydrological Model

### 2015-06-17 TeXFiles/stochInf1D.tex
update of tex files and few other things. Issue with sin rain not fixed yet

According to \cite{lau_2007_StateDepDiff}, the probability, $P(q_1,T|q_0,0)$, of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0) 
= 
\frac{1}{Z}
\int 
e^{-{\mathcal S}[q,\dot q]} 
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as an infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as
%\begin{equation}\label{action}
%  \tilde{\mathcal S}[\dot{S}(t),S(t)]
%  =
%  \int_0^T
%  dt
%  \left\{
%    \frac{K}{2\gamma S^2(t)}
%    \left(
%        \dot S(t)
%        -
%        r(t)
%        +
%        \frac{1+\gamma}{K}S(t)
%    \right)^2
%    -
%    \frac{2+\gamma}{4K}
%  \right\}
% \,.
%\end{equation}
\begin{equation}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite action (\ref{action}) as,
\begin{multline}\label{action}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0)
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{multline}

@@

The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
For the time being we assume constant priors.
Then, the posterior reads as
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y, \vc r)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[{q},\dot q]
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r_s)-{\beta q_s})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
with
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m_q}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f(\bt | \vc y, \vc r)
  \propto
  \int
  e^{-\mathcal {H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
  \wedge
  {\mathcal Dq}
  \wedge
  d{\pmb\pi}
  \,.
\end{equation}

@@

Let us first write down the discretization of the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
Path-integral (\ref{posterior_pathint}) is approximated by an ordinary integral, with the approximate path-measure 
\begin{equation}
  \mathcal Dp\wedge\mathcal Dq
  \approx
  \prod_i dp_i\wedge dq_i 
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{align}
   K( p,{\pmb\pi})
   &\approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt)
   &\approx
   \frac{\Delta t}{T}
   \sum_{i=2}^{N}
   \left\{
    \frac{1}{2} 
    T^2 \dot q_i^2
    +
    \frac{1}{2}
     \left(
        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
     \right)^2
    -
    \frac{\beta^2}{2\gamma}
    e^{-\beta q_i}
   -
    T q_i\dot\rho_i
   \right\}
  \\
  &+
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  & +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
   \,,\label{Vdisc}
\end{align}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r_{i+1}/r_i)}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}) and that we've assumed $r_i$ to be the discretization of an at least twice differentiable interpolation of the input measurements.

@@

We choose another approach, the so-called \emph{staging} as described in \cite{tuckerman1992reversible}, and diagonalize the harmonic part only in between the measurement points.
We first note that Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.

The discretized harmonic part of the action is rewritten as
\begin{multline}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2
  \\ =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,
\end{multline}
with
\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}
Following \cite{tuckerman1993efficient} we apply the coordinate transformation
\begin{equation}
  u_{sj+1} = q_{sj+1}\,,\quad
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
\end{equation}
with its inverse given by
\begin{equation}
  q_{sj+1} = u_{sj+1}\,,\quad
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{equation}
or equivalently by the recursive relation
\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{s+k+1}+ \frac{1}{k}q_{sj+1} \,.
\end{equation}
%Moreover, we introduce the following transformation for the parameters,
%
%\begin{equation}\label{params_transform}
%\beta = \ln(K/T) \,,\quad \tau=\ln(\gamma)
%\end{equation}
%
%and the corresponding transformed vector $\tilde{\bt}=(\beta,\tau)^T$. This transformation ensures that the system parameters $K$ and $\gamma$ will be kept positive during the HMC algorithm.
%

This transformation decouples fast degrees of freedom from slow ones and allows us to split the Hamiltonian $\mathcal H_{HMC}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$.
We write
\begin{equation}
  \mathcal H_{HMC}= \mathcal H_N+\mathcal H_n+\mathcal H_1\,,
\end{equation}
where
\begin{align}
  \mathcal H_N
  &=
  \frac{1}{2}
  \sum_{s=1}^{n}
  \sum_{k=2}^j
  \left\{
    \frac{\Delta t}{m_q}p_{(s-1)j+k}^2
    +
    \frac{Tk}{\Delta t(k-1)}
    u_{(s-1)j+k}^2
  \right\}\,,\label{H_N}
  \\
  \mathcal H_n
  &=
  \frac{1}{2}
  \sum_{s=1}^{n+1}
  \left\{
   \frac{\Delta t }{m_q}p_{(s-1)j+1}^2
    +
    \frac{T}{j\Delta t}
    (u_{(s-1)j+1} - u_{sj+1})^2
    -
    \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta u_{(s-1)j+1}})^2}{2\sigma^2}
   \right\}\,,\\
  \mathcal H_1
  &=
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}
   +
  \frac{\Delta t}{T}
   \sum_{i=2}^{N}
   \left\{
    \frac{1}{2}
     \left(
        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
     \right)^2
    -
    \frac{\beta^2}{2\gamma}
    e^{-\beta q_i}
   -
    T q_i\dot\rho_i
   \right\}
  \\
  &+
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
\,.
\end{align}
The harmonic part $\mathcal H_N$, for the staging beads, scales linearly with $N$.
The harmonic term $\mathcal H_n$, for the boundary beads, as well as the measurement term scale linearly with $n$.
Finally, the potential does not scale with neither $n$ nor $N$.
Thus, depending on the values of $N$ and $n$, the dynamics happens on very different time scales.
Therefore, we use the Trotter formula according to \cite{tuckerman1992reversible} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes.
\begin{enumerate}
  \item[\it i.]
  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$
  \item[\it ii.]
  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$
  \item[\it iii.]
  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$
\end{enumerate}

In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
\begin{equation}
  \mathcal H_{HMC}=\mathcal H_N + \mathcal H'\,.
\end{equation}
In order to design a reversible integrator, for the associated Hamiltonian equations, we define the Liouville operators
\begin{equation}
  iL_N=\{\cdot\,,\,\mathcal H_N\}\,,\quad
  iL'=\{\cdot\,,\,\mathcal H'\}\,,
\end{equation}
where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
Trotter's formula allows us to write the Hamiltonian propagator as
\begin{equation}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(P^{-2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes a much faster dynamics than the inner one.
However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m_q/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The fast outer propagator is then given by eqs.
\begin{align}
  u_{(s-1)j+k}(\Delta\tau/2)
  &=
  u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
  p_{(s-1)j+k}(\Delta\tau/2)
  &=
  p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{align}
for $k=2,\dots,j$.

>>>>

According to \cite{lau_2007_StateDepDiff}, the probability, $P(q_1,T|q_0,0)$, of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as an infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as
%\begin{equation}\label{action}
%  \tilde{\mathcal S}[\dot{S}(t),S(t)]
%  =
%  \int_0^T
%  dt
%  \left\{
%    \frac{K}{2\gamma S^2(t)}
%    \left(
%        \dot S(t)
%        -
%        r(t)
%        +
%        \frac{1+\gamma}{K}S(t)
%    \right)^2
%    -
%    \frac{2+\gamma}{4K}
%  \right\}
% \,.
%\end{equation}
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite action (\ref{action1}) as,
\begin{multline}\label{action}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0)
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{multline}

@@

The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
For the time being we assume constant priors.
Then, the posterior reads as
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y, \vc r)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[{q},\dot q]
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r_s)-{\beta q_s})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
with
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f(\bt | \vc y, \vc r)
  \propto
  \int
  e^{-\mathcal {H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
  \wedge
  {\mathcal Dq}
  \wedge
  d{\pmb\pi}
  \,.
\end{equation}

@@

Let us first write down the discretization of the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
Path-integral (\ref{posterior_pathint}) is approximated by an ordinary integral, with the approximate path-measure
\begin{equation}
  \mathcal Dp\wedge\mathcal Dq
  \approx
  \prod_i dp_i\wedge dq_i
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{align}
   K( p,{\pmb\pi})
   &\approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt)
   &\approx
   \frac{\Delta t}{T}
   \sum_{i=2}^{N}
   \left\{
    \frac{1}{2}
    T^2 \dot q_i^2
    +
    \frac{1}{2}
     \left(
        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
     \right)^2
    -
    \frac{\beta^2}{2\gamma}
    e^{-\beta q_i}
   -
    T q_i\dot\rho_i
   \right\}\label{Vdisc}
  \\
  &+
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  & +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  \nonumber
   \,,
\end{align}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r_{i}/r_{i-1})}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}) and that we've assumed $r_i$ to be the discretization of an at least twice differentiable interpolation of the input measurements.

@@

We choose another approach, the so-called \emph{staging} as described in \cite{tuckerman1992reversible}, and diagonalize the harmonic part only in between the measurement points.
We first note that the Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.

The discretized harmonic part of the action is rewritten as
\begin{multline}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2
  \\ =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,
\end{multline}
with
\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}
Following \cite{tuckerman1993efficient} we apply the coordinate transformation
\begin{equation}
  u_{sj+1} = q_{sj+1}\,,\quad
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
\end{equation}
with its inverse given by
\begin{equation}
  q_{sj+1} = u_{sj+1}\,,\quad
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{equation}
or equivalently by the recursive relation
\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{sj+k+1}+ \frac{1}{k}u_{sj+1} \,.
\end{equation}
%Moreover, we introduce the following transformation for the parameters,
%
%\begin{equation}\label{params_transform}
%\beta = \ln(K/T) \,,\quad \tau=\ln(\gamma)
%\end{equation}
%
%and the corresponding transformed vector $\tilde{\bt}=(\beta,\tau)^T$. This transformation ensures that the system parameters $K$ and $\gamma$ will be kept positive during the HMC algorithm.
%

This transformation decouples fast degrees of freedom from slow ones and allows us to split the Hamiltonian $\mathcal H_{HMC}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$.
We write
\begin{equation}
  \mathcal H_{HMC}= \mathcal H_N+\mathcal H_n+\mathcal H_1\,,
\end{equation}
where
\begin{align}
  \mathcal H_N
  &=
  \frac{1}{2}
  \sum_{s=1}^{n}
  \sum_{k=2}^j
  \left\{
    \frac{\Delta t}{m_q}p_{(s-1)j+k}^2
    +
    \frac{Tk}{\Delta t(k-1)}
    u_{(s-1)j+k}^2
  \right\}\,,\label{H_N}
  \\
  \mathcal H_n
  &=
  \frac{1}{2}
  \sum_{s=1}^{n+1}
  \left\{
   \frac{\Delta t }{m_q}p_{(s-1)j+1}^2
    +
    \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta u_{(s-1)j+1}})^2}{\sigma^2}
   \right\}
   \\
  \nonumber
  &+
  \frac{T}{2j\Delta t}
  \sum_{s=1}^{n}
    (u_{(s-1)j+1} - u_{sj+1})^2
   \,,\\
  \mathcal H_1
  &=
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}
   +
  \frac{\Delta t}{T}
   \sum_{i=2}^{N}
   \left\{
    \frac{1}{2}
     \left(
        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
     \right)^2
    -
    \frac{\beta^2}{2\gamma}
    e^{-\beta q_i}
   -
    T q_i\dot\rho_i
   \right\}
  \\
  \nonumber
  &+
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
\,.
\end{align}
The harmonic part $\mathcal H_N$, for the staging beads, scales linearly with $N$.
The harmonic term $\mathcal H_n$, for the boundary beads, as well as the measurement term scale linearly with $n$.
Finally, the potential does not scale with neither $n$ nor $N$.
Thus, depending on the values of $N$ and $n$, the dynamics happens on very different time scales.
Therefore, we use the Trotter formula according to \cite{tuckerman1992reversible} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes.
\begin{enumerate}
  \item[\it i.]
  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$
  \item[\it ii.]
  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$
  \item[\it iii.]
  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$
\end{enumerate}

In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
\begin{equation}
  \mathcal H_{HMC}=\mathcal H_N + \mathcal H'\,.
\end{equation}
In order to design a reversible integrator, for the associated Hamiltonian equations, we define the Liouville operators
\begin{equation}
  iL_N=\{\cdot\,,\,\mathcal H_N\}\,,\quad
  iL'=\{\cdot\,,\,\mathcal H'\}\,,
\end{equation}
where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
Trotter's formula allows us to write the Hamiltonian propagator as
\begin{equation}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(P^{-2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes a much faster dynamics than the inner one.
However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m_q/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The fast outer propagator is then given by eqs.
\begin{align}
  u_{(s-1)j+k}(\Delta\tau/2)
  &=
  u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
  p_{(s-1)j+k}(\Delta\tau/2)
  &=
  p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{align}
for $s=1,\dots,n$ and $k=2,\dots,j$.
